Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-09-12 paper.tex
Finished first draft of introduction.

Type Ia supernovae (SNe Ia) are currently some of the most exciting events to study in astrophysics. These bright, brief pulses of light in the distant universe have led to a number of important discoveries in recent years, including the discovery of the accelerated expansion of the universe \citep{perlmutter1999,riess1998}. However, their origin is shrouded in mystery. It has long been expected that these events arise from the thermonuclear explosions of white dwarfs \citep{hoyle_fowler:1960}, but the cause of these explosions is uncertain. In particular, it is not clear what process causes the temperatures in these white dwarfs to become hot enough for explosive burning of its constituent nuclei. The model favored initially by the community was the so-called single-degenerate (SD) model. Accretion of material from a companion star such as a red giant would cause the star to approach the Chandrasekhar mass, and in doing so the temperature and density in the center would become sufficient for thermonuclear fusion to proceed. A review of the observational evidence and theoretical modelling that formed this belief can be found in \citet{hillebrandtniemeyer2000}.

% History of single-degenerate model (one paragraph)

% Observational evidence that disfavors SD model (one paragraph)

% Early theoretical history of double-degenerate model (one paragraph)

% Observational evidence that favors DD model (two-three paragraphs)

@@

The most recent developments give some areas of agreement and some of remaining uncertainty. \cite{pakmor:2012} performed a merger scenario with a $1.1\ M_\odot$ and $0.9\ M_\odot$ setup, with better treatment of the initial conditions, and also found that the merger process happened over more than ten orbits. Nevertheless, they still found that a carbon-oxygen detonation would occur, in line with their earlier results. \cite{moll:2014} was also able to find a detonation in a similarly massive system. \cite{dan:2012} and \cite{dan:2014} performed a large sweep of the parameter space for merger pairs and found that pure carbon-oxygen systems would generally not lead to detonations and violent mergers except for the most massive systems. They did find that for systems containing helium, many of them would detonate and potentially lead to SNe Ia, either through the aforementioned instabilities in the accretion stream, or during the contact phase, similar to the violent carbon-oxygen mergers. \cite{pakmor:2013} added a thin helium shell on their primary white dwarf, and found that this robustly led to a detonation of the white dwarf. Thus the preliminary consensus seems to be that systems containing helium could robustly lead to events resembling SNe Ia, as well as very massive carbon-oxygen binaries.

% Still need Raskin/Moll/et al. papers

% Motivation: a series of papers to verify or disprove the prompt/violent merger hypothesis using a grid-based code

@@

%==========================================================================
% Test Problems
%==========================================================================
\section{Test Problems}\label{Sec:Tests}

@@

%==========================================================================
% Performance
%==========================================================================
\section{Performance}\label{Sec:Performance}

@@

%==========================================================================
% Conclusions
%==========================================================================
\section{Conclusions and Discussion}\label{Sec:Conclusions and Discussion}

>>>>

Type Ia supernovae (SNe Ia) are currently some of the most exciting events to study in astrophysics. These bright, brief pulses of light in the distant universe have led to a number of important discoveries in recent years, including the discovery of the accelerated expansion of the universe \citep{perlmutter1999,riess1998}. However, their origin is shrouded in mystery. It has long been expected that these events arise from the thermonuclear explosions of white dwarfs \citep{hoyle_fowler:1960}, but the cause of these explosions is uncertain. In particular, it is not clear what process causes the temperatures in these white dwarfs to become hot enough for explosive burning of its constituent nuclei. The model favored initially by the community was the so-called single-degenerate (SD) model \citep{whelan_iben:1973}. Accretion of material from a companion star such as a red giant would cause the star to approach the Chandrasekhar mass, and in doing so the temperature and density in the center would become sufficient for thermonuclear fusion to proceed. However, in recent years many alternative progenitor models have been discussed. A leading candidate for explaining some or most of these explosions is the double-degenerate model, in which two white dwarfs merge and the merged object reaches the conditions necessary for a thermonuclear ignition \citep{ibentutukov:1984,webbink:1984}. Another is the double detonation scenario, where accretion of material onto a sub-Chandrasekhar white dwarf would lead to a detonation inside the accreted envelope, and this would send a compressional wave into the core of the star that would trigger a secondary detonation. A recent review of the progenitor models can be found in \citet{hillebrandt:2013}.

There are several observational reasons why double-degenerate systems are a promising progenitor system for at least a substantial fraction of normal SNe Ia. No conclusive evidence exists for a surviving companion star of a SN Ia; this is naturally explained by the DD model because both WDs are destroyed in the merger process. Similarly, pre-explosion images of the SN Ia systems have never clearly turned up a companion star, and in some cases a large fraction of the SD parameter space is excluded. Additionally, not enough progenitor systems are seen for the SD case to match the observed local SN Ia rate, whereas the number of white dwarf binaries may be sufficient to account for this rate. Finally, the DD model can naturally explain the fact that many SNe Ia are observed to occur at very long delay times after the stars were formed, since the progenitor systems only become active once both stars have evolved off the main sequence. A thorough review of the observational evidence about SNe Ia and further discussion of these ideas can be found in \cite{maoz:2014}.

@@

The most recent developments give some areas of agreement and some of remaining uncertainty. \cite{pakmor:2012} performed a merger scenario with a $1.1\ M_\odot$ and $0.9\ M_\odot$ setup, with better treatment of the initial conditions, and also found that the merger process happened over more than ten orbits. Nevertheless, they still found that a carbon-oxygen detonation would occur, in line with their earlier results. \cite{moll:2014} was also able to find a detonation in a similarly massive system. \cite{dan:2012} and \cite{dan:2014} performed a large sweep of the parameter space for merger pairs and found that pure carbon-oxygen systems would generally not lead to detonations and violent mergers except for the most massive systems. They did find that for systems containing helium, many of them would detonate and potentially lead to SNe Ia, either through the aforementioned instabilities in the accretion stream, or during the contact phase, similar to the violent carbon-oxygen mergers. \cite{pakmor:2013} added a thin helium shell on their primary white dwarf, and found that this robustly led to a detonation of the white dwarf. Thus at least a preliminary agreement may be that systems containing helium could robustly lead to events resembling SNe Ia, as well as very massive carbon-oxygen binaries.

This is the first in a series of papers designed to address these outstanding theoretical issues for white dwarf mergers. This work will discuss the verification of our hydrodynamics code for simulating these events. Later efforts will look at the initial conditions of the system, the robustness with which a hotspot is found from which a detonation could occur, and the importance of the initial white dwarf models, which should be more sophisticated than simple carbon-oxygen mixtures and in principle should use results from modern stellar evolution calculations. Section \ref{sec:Numerical Methodology} describes our code and why it can provide useful results compared to other methodologies used for this problem. Section \ref{sec:Tests} discusses a few test problems that we use to demonstrate that our code accurately solves the equations of fluid dynamics. Section \ref{sec:Performance} demonstrates that the software scales well for supercomputer applications. Finally, Section \ref{sec:Conclusions and Discussion} recaps what we have shown and highlights some of the future work we plan to do.

@@

%==========================================================================
% Test Problems
%==========================================================================
\section{Test Problems}\label{sec:Tests}

@@

%==========================================================================
% Performance
%==========================================================================
\section{Performance}\label{sec:Performance}

@@

%==========================================================================
% Conclusions
%==========================================================================
\section{Conclusions and Discussion}\label{sec:Conclusions and Discussion}
